Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-10-09 content.tex
Incorporating DVH's comments.

%%
A downside of the work providing \rackett{call/cc} is that it was an algorithmic change to the already complicated CFA2 --- there was no recipe for how to do this for one's favorite control operator.
%
This paper seeks to do just that with an operational view of what summarization is, in essence.
%
In other words, we give a concrete semantics to the tricks that the analyses in the literature use in the abstract, and maintain the original meaning of the language.
%
We also give intuitive analogies to well-established ideas and techniques so that the working semanticist can write a pushdown analysis for her language.
%
In order to demonstrate the applicability of this viewpoint, we show a new analysis for a language with composable control.
%
All of the semantics modeled in this paper are implemented in full detail in PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available online\footnote{\url{http://github.com/ianj/concrete-summaries}}.

@@

To turn this semantics into PDCFA's algorithm for constructing a Dyck state graph, given the appropriate $\alloc$, we apply a widening operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these components), in what is then called a $\System$. We use a meta-function $\wn$, ``wide to narrow,'' to shift between the two representations of states.
%%

%%
We show only the memo and continuation table widening, since store widening would require rewriting newly-created contexts to use the global store following the step.
%
This rewriting is unnecessary with the store-counting per-wide-step technique presented in \citet{ianjohnson:oaam:icfp2013}, or the less precise timestamp approximation from \citet{dvanhorn:Shivers:1991:CFA}.
%
Since these components are shared, they are always the least upper bound of any respective component that the semantics produces (the intermediate set $I$) in order to stay sound.

@@

%%
We should expect that a pushdown analysis would predict this evaluates to true, and there are no loops in the program.
%
0CFA~\citep{dvanhorn:Shivers:1991:CFA} claims there is a loop from the second call of \texttt{id} to the first, and thusly predicts this program evaluates to true or false.
%
PDCFA claims there is no loop, and depending on the implementation, that the result is true or false (paper), or just true (implemented)
\footnote{This is because the paper steps every seen state with the current store every iteration, but the implementation only steps states that need stepping.}.
%
The example evaluation is in \autoref{fig:ex-eval}.
%
We maintain enough context to distinguish the return points of \texttt{id} to not rebind \texttt{y} to 1.
%
When determining control flow through the expression, we consult $\mmemo$ to continue past function calls, so there is no confusion about a back edge from the second call to \texttt{id}.
%

@@

\begin{theorem}\label{thm:pdcfa-tabular}
  Tabular semantics has a bisimulation with the standard semantics with comparable allocation strategies.
\end{theorem}

This follows directly from the invariants we prove of $\mktab$ and $\mmemo$ in the tabular semantics.
%
Allocation strategies are comparable if they produce equal addresses regardless of the differences in the state representations.
%
More formally, $\alloc$ and $\alloc^*$ are comparable if this implication holds:
\begin{mathpar}
  \inferrule{\mkont \in \unroll{\mktab}{\mkont^*}}
            {\alloc(\tpl{\mpoint,\mstore,\mkont}) = \alloc^*(\tpl{\mpoint,\mstore,\mkont^*,\mktab,\mmemo})}
\end{mathpar}

The $\mathit{unroll}$ function interprets what are all the valid continuations that $\mktab$ encodes for a given continuation that contains $\krt{\mctx}$, defined as the greatest fixed point of the following rules:
\begin{mathpar}
  \inferrule{ }{\kmt \in \unroll{\mktab}{\kmt, G}} \quad
  \inferrule{\mkont \in \unroll{\mktab}{\mkont', \emptyset}}{\mkframe:\mkont \in \unroll{\mktab}{\mkframe:\mkont', G}} \\
  \inferrule{\mctx \notin G  \\
             \mkont^\circ \in \mktab(\mctx) \\
             \mkont \in \unroll{\mktab}{\mkont^\circ, \set{\mctx}\cup G}}
            {\mkont \in \unroll{\mktab}{\krt{\mctx}, G}}  
\end{mathpar}

@@

\begin{figure}
  \centering
  \begin{align*}
    \inv(\tpl{\mpoint, \mstore, \mkont, \mktab, \mmemo}) &=
    \dom(\mmemo) \subseteq \dom(\mktab) \\
    &\wedge \forall (\mpoint',\mstore') \in \dom(\mktab), \mkont^\circ
    \in \mktab(\mpoint',\mstore'),
    \mkont' \in \unroll{\mktab}{\mkont^\circ, \emptyset}. \\
    &\qquad \phi_{\reachable} \wedge \phi_{\memo} \\
    \text{where } \phi_{\reachable} &= \exists \mkont^* \in
    \unroll{\mktab}{\mkont,\emptyset}.
    \tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mpoint, \mstore, \mkont^*} \\
    \phi_{\memo} &=
    \forall (\mval, \mstore'') \in \mmemo(\mpoint',\mstore'). \\
    &\qquad \exists \mtrace\equiv\tpl{\mpoint', \mstore', \mkont'}
    \stepto^* \tpl{\mval,\mstore'', \mkont'}.
    \hastail(\mtrace,\mkont') \\
  \end{align*}
  \caption{Table invariants}
\label{fig:inv}
\end{figure}
\begin{lemma}[Table invariants]\label{lem:tab-inv}
  If $\inv(\mstate)$ and $\mstate \stepto \mstate'$ then $\inv(\mstate')$.
\end{lemma}

In order to prove this, we need a lemma that allows us to plug in any continuation for memoized results.

\begin{lemma}[Stack irrelevance]\label{lem:irrelevance}
  $\hastail(\mtrace,\mkont)$ implies $\replacetail(\mtrace,\mkont,\mkont')$ is a valid trace.
\end{lemma}
\begin{mathpar}
  \inferrule{ }{\hastail(\epsilon,\mkont)} \quad
  \inferrule{ }{\hastail(\tpl{\mpoint,\mstore,\append{\mkont'}{\mkont}},\mkont)} \quad
  \inferrule{\hastail(\mtrace\mstate,\mkont) \quad
             \mstate \stepto \mstate' \quad
             \hastail(\mstate',\mkont)}
            {\hastail(\mtrace\mstate\mstate',\mkont)}
\end{mathpar}

@@

Finally, we can show that even if the tables are shared amongst all states, they are precise enough to not interfere with the results.
\begin{theorem}[Equivalence with global tables]\label{thm:global}
  Tabular semantics has a bisimulation with memo-and-k-table-widened semantics.
\end{theorem}

@@

An implementation of the same functionality would use a mutable set to track which contexts have been traversed to build up a set of addresses:
% This code is typeset below by Scribble
% \begin{lstlisting}[mathescape]
% (define (touch-kont $\mkont$ $\mktab$)
%   (define seen (mutable-seteq))
%   (let loop ([$\mkont$ $\mkont$])
%     (match $\mkont$
%       [(mt) $\emptyset$]
%       [(cons $\mkframe$ $\mkont$) ($\cup$ $\touchesf(\mkframe)$ (loop $\mkont$))]
%       [(rt $\mctx$)
%        (cond
%          [(set-member? seen $\mctx$) $\emptyset$]
%          [else
%           (set-add! seen $\mctx$)
%           (for/union ([$\mkont$ (in-set $\mktab(\mctx)$)]) (loop $\mkont$))])])))           
% \end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Begin scary Scribble output
\begin{SCodeFlow}\begin{RktBlk}\begin{SingleColumn}\RktPn{(}\RktSym{define}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{touch{-}kont}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\kappa$}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\Xi$}\RktPn{)}

\mbox{\hphantom{\Scribtexttt{xx}}}\RktPn{(}\RktSym{define}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{seen}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{mutable{-}seteq}\RktPn{)}\RktPn{)}

\mbox{\hphantom{\Scribtexttt{xx}}}\RktPn{(}\RktSym{define}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{fix{-}touch}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\kappa$}\RktPn{)}

\mbox{\hphantom{\Scribtexttt{xxxx}}}\RktPn{(}\RktSym{match}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\kappa$}

\mbox{\hphantom{\Scribtexttt{xxxxxx}}}\RktPn{[}\RktPn{(}\RktSym{mt}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\emptyset$}\RktPn{]}

\mbox{\hphantom{\Scribtexttt{xxxxxx}}}\RktPn{[}\RktPn{(}\RktSym{cons}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\phi$}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\kappa$}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{$\cup$}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{touches{-}frame}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\phi$}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{fix{-}touch}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\kappa$}\RktPn{)}\RktPn{)}\RktPn{]}

\mbox{\hphantom{\Scribtexttt{xxxxxx}}}\RktPn{[}\RktPn{(}\RktSym{rt}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{ctx}\RktPn{)}

\mbox{\hphantom{\Scribtexttt{xxxxxxx}}}\RktPn{(}\RktSym{cond}

\mbox{\hphantom{\Scribtexttt{xxxxxxxx}}}\RktPn{[}\RktPn{(}\RktSym{set{-}member{\hbox{\texttt{?}}}}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{seen}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{ctx}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\emptyset$}\RktPn{]}

\mbox{\hphantom{\Scribtexttt{xxxxxxxx}}}\RktPn{[}\RktSym{else}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{set{-}add{\hbox{\texttt{!}}}}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{seen}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{ctx}\RktPn{)}

\mbox{\hphantom{\Scribtexttt{xxxxxxxxxxxxxx}}}\RktPn{(}\RktSym{union{-}map}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{fix{-}touch}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{ref}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\Xi$}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{ctx}\RktPn{)}\RktPn{)}\RktPn{]}\RktPn{)}\RktPn{]}\RktPn{)}\RktPn{)}

\mbox{\hphantom{\Scribtexttt{xx}}}\RktPn{(}\RktSym{fix{-}touch}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{$\kappa$}\RktPn{)}\RktPn{)}\end{SingleColumn}\end{RktBlk}\end{SCodeFlow}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% End of scary Scribble output

@@

This semantics of course enjoys the soundness properties that GC does not introduce dangling pointers (\autoref{thm:gc-pointers}), equal length traces are equal up to GC (in the concrete) (\autoref{thm:gc-concrete}), and abstractions are sound with respect to collected states (\autoref{thm:gc-sound}).
%
Technically since this semantics inspects the touched addresses of the continuation, the set of touched addresses should be considered part of the ``context'' in the $\mathbf{rt}$ continuations.
%
This extra context can be overlooked, since that means the computed touched addresses will be an overapproximation of what is already sound.
%
More touched addresses means a less effective GC, but this is an acceptable tradeoff for a possibly smaller state space.
%
\begin{theorem}[No dangling pointers]\label{thm:gc-pointers}
  All addresses in $\Gamma(\mstate)$ are in the domain of $\Gamma(\mstate).\mstore$.
\end{theorem}

\begin{theorem}[Garbage is irrelevant]\label{thm:gc-concrete}
  If $\mstate \stepto^n \mstate'$ and $\mstate \stepto^n \mstate''$ then $\Gamma(\mstate') = \Gamma(\mstate'')$.
\end{theorem}

\begin{theorem}[Abstract GC is sound for garbage-free states]\label{thm:gc-sound}
  If $\mstate \stepto \mstate'$ and $\alpha(\Gamma(\mstate)) \sqsubseteq \mastate$ then there is a $\mastate'$ such that $\mastate \astepto \mastate'$ and $\alpha(\Gamma(\mstate')) \sqsubseteq \mastate'$.
\end{theorem}

This final theorem is not in the original work on abstract GC; they prove soundness with respect to a concrete semantics that performs GC after every step.
%
The moral correctness of the abstraction is that it never under-approximates \emph{relevant} portions of the state space --- in this case, the reachable subset of the store.
%
The definitions of $\alpha$, $\astepto$ and $\sqsubseteq$ are the simple pointwise abstractions of addresses through states and the reduction relation, and finally a lifted subset relation between states and their components, respectively.
%
Both $\alpha$ and $\sqsubseteq$ are parameterized on the choice of $\Addr$, for which the following soundness criterion should hold:
\begin{mathpar}
  \inferrule*[right={[Sound allocation]}]{\alpha(\mstate) \sqsubseteq \mastate}{\alpha(\alloc(\mstate)) \sqsubseteq \widehat{\alloc}(\mastate)}
\end{mathpar}
\section{Deriving CFA2}
\label{sec:cfa2}

CFA2 is the first published analysis of a higher-order programming language that could properly match calls and returns.
%
We will show that it fits well into the same presentation we gave for PDCFA.
%
Vardoulakis and Shivers had a clear goal of harnessing the extra information a pushdown model provides to produce a high-precision analysis that works well in practice.
%
This resulted in more than just the call/return matching of the previous section, which is why we are showing the two separately.
%
There are two orthogonal features of the semantics, embodied by our semantics' two metafunctions ($\bind$ and $\lookup$, respectively in \autoref{fig:frame-semantics}):
\begin{enumerate}
\item{stack allocation for some bindings in an additional $\mframe \in \Store$}
\item{strong updates on stack frames for resolved non-determinism}
\end{enumerate}
The first of these is an addition to the stack-less context.

@@

%%
The second of these is to ameliorate a problem they call ``fake rebinding.''
%
That is, since bindings in the abstract represent several values, we don't want to reference a variable \texttt{x} in two different places and have it resolve to two different values.
%
In AAM, a variable reference non-deterministically steps to all possible values associated with that variable.
%
Here we want to say that once \texttt{x} is considered to stand for value $v$, then all subsequent references of \texttt{x} should be $v$.
%
If they are not all considered $v$, it looks as if \texttt{x} was rebound; it hasn't, and thus it is a ``fake rebinding.''
%
CFA2 does not step to all values on variable reference, but instead carries all its values around in superposition until they need to be observed at, say, a function call.
%
We give a simplified semantics that is more along the AAM style, but CFA2's approach can easily be recovered from it\footnote{Our Redex model implements the fake rebinding strategy CFA2 itself employs}.
%%

@@

\begin{align*}
  \bind(\mstore,\maddr,\mvar,\mval) &=
   \left\{\begin{array}{l}
            (\mstore, \snglm{\maddr}{\mval}) \text{ if } \mvar \text{ allocations never escape} \\
            (\joinone{\mstore}{\maddr}{\mval}, \snglm{\maddr}{\mval}) \text{ otherwise} \\
          \end{array}\right. \\
  \lookup(\mstore,\mframe,\menv,\mvar,\mlab) &=
    \left\{\begin{array}{l}
          \setbuild{(\extm{\mframe}{\menv(\mvar)}{\set{\mval}}, \mval)}
                   {\mval \in \mframe(\menv(\mvar))}
                   \\ \qquad \text{if } \mlab \text{ non-escaping} \\
          \setbuild{(\mframe,\mval)}{\mval \in \mstore(\menv(\mvar))} \text{ otherwise}
           \end{array}\right.
\end{align*}

@@

%%
The stack frame introduces a refinement over the original CESK machine, so we first prove it correct with respect to the concrete semantics.
%
We follow up with a similar proof to that of PDCFA, which is that the CES$\xi$K machine has a bisimulation with its tabular semantics, with comparable allocation strategies.
\begin{theorem}[Correctness of stack frame refinement]\label{thm:refinement}
  The CES$\xi$K machine has a bisimulation with the CESK machine with fresh allocation.
\end{theorem}
This follows from the invariant that if a stack frame maps $\maddr$ to $\mval$, then the store maps $\maddr$ to $\mval$, meaning that the special lookup function is equivalent to lookup directly in the store (exploits immutability).
%
That is, if we always allocated bindings in the heap.
%
The additional escape information that allows us to elide allocation in the heap introduces a second level of reasoning, where the burden of proof falls on the escape analysis.
%
The additional escape analysis adds information to use in the proof, that if $\mvar$ never escapes, then every reference to allocations of $\mvar$'s binding will always be able to use $\mframe$ for the lookup.
%
With this proposition, we can prove that the CES$\xi$K machine that always allocates in the store has a bisimulation with the refined machine that exploits escape analysis to avoid unnecessary changes to the store.
%

\begin{theorem}[Correctness of CFA2]\label{thm:cfa2}
  Tabular semantics has a bisimulation with standard semantics.
\end{theorem}
This proof has the same structure as that of PDCFA.

@@

\begin{theorem}\label{thm:concrete-sr}
  Tabular semantics has a bisimulation with standard semantics given fresh allocation.
\end{theorem}

@@

\begin{theorem}\label{thm:sound-sr}
  Tabular semantics simulates standard semantics with abstract allocation.
\end{theorem}

>>>>

%%
A downside of the work providing \rackett{call/cc} is that it was an algorithmic change to the already complicated CFA2 --- there was no recipe for how to do this for one's favorite control operator.
%
This paper seeks to do just that with an operational view of what summarization is, in essence.
%
In other words, we give a concrete semantics to the tricks that the analyses in the literature use in the abstract, and maintain the original meaning of the language.
%
We also give intuitive analogies to well-established ideas and techniques so that the working semanticist can write a pushdown analysis for her language.
%
In order to demonstrate the applicability of this viewpoint, we show a new analysis for a language with composable control.
%
All of the semantics modeled in this paper are implemented in full detail in PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available online.\footnote{\url{http://github.com/ianj/concrete-summaries}}

@@

To turn this semantics into PDCFA's algorithm for constructing a Dyck state graph, given the appropriate $\alloc$, we apply a widening operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these components), in what is then called a $\System$. We use a meta-function $\wn$, ``wide to narrow,'' to shift between the two representations of states.
%
We show only the memo and continuation table widening, since store widening is standard.
%

@@

%%
We should expect that a pushdown analysis would predict this evaluates to true, and there are no loops in the program.
%
0CFA~\citep{dvanhorn:Shivers:1991:CFA} claims there is a loop from the second call of \texttt{id} to the first, and thusly predicts this program evaluates to true or false.
%
PDCFA claims there is no loop, and depending on the implementation, that the result is true or false (paper), or just true (implemented).
\footnote{This is because the paper steps every seen state with the current store every iteration, but the implementation only steps states that need stepping.}
%
The example evaluation is in \autoref{fig:ex-eval}.
%
We maintain enough context to distinguish the return points of \texttt{id} to not rebind \texttt{y} to 1.
%
When determining control flow through the expression, we consult $\mmemo$ to continue past function calls, so there is no confusion about a back edge from the second call to \texttt{id}.
%

@@

\subsection{Metatheory}

Correctness follows directly from the invariants we prove of $\mktab$ and $\mmemo$ in the tabular semantics.
%
Allocation strategies are comparable if they produce equal addresses regardless of the differences in the state representations.
%
More formally, $\alloc$ and $\alloc^*$ are comparable if this implication holds:
\begin{mathpar}
  \inferrule{\mkont \in \unroll{\mktab}{\mkont'}}
            {\alloc(\tpl{\mpoint,\mstore,\mkont}) = \alloc^*(\tpl{\mpoint,\mstore,\mkont',\mktab,\mmemo})}
\end{mathpar}

The $\mathit{unroll}$ function interprets what are all the valid continuations that $\mktab$ encodes for a given continuation that contains $\krt{\mctx}$, defined as the greatest fixed point of the following rules:
\begin{mathpar}
  \inferrule{ }{\kmt \in \unroll{\mktab}{\kmt, G}} \quad
  \inferrule{\mkont \in \unroll{\mktab}{\mkont', \emptyset}}{\mkframe:\mkont \in \unroll{\mktab}{\mkframe:\mkont', G}} \\
  \inferrule{\mctx \notin G  \\
             \mkont'' \in \mktab(\mctx) \\
             \mkont \in \unroll{\mktab}{\mkont'', \set{\mctx}\cup G}}
            {\mkont \in \unroll{\mktab}{\krt{\mctx}, G}}  
\end{mathpar}

@@

\begin{figure}
  \centering
  \begin{align*}
    \inv(\tpl{\mpoint, \mstore, \mkont, \mktab, \mmemo}) &=
    \dom(\mmemo) \subseteq \dom(\mktab) \\
    &\wedge \forall (\mpoint',\mstore') \in \dom(\mktab), \mkont''
    \in \mktab(\mpoint',\mstore'),
    \mkont' \in \unroll{\mktab}{\mkont'', \emptyset}. \\
    &\qquad \phi_{\reachable} \wedge \phi_{\memo} \\
    \text{where } \phi_{\reachable} &= \exists \mkont'' \in
    \unroll{\mktab}{\mkont,\emptyset}.
    \tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mpoint, \mstore, \mkont''} \\
    \phi_{\memo} &=
    \forall (\mval, \mstore'') \in \mmemo(\mpoint',\mstore'). \\
    &\qquad \exists \mtrace\equiv\tpl{\mpoint', \mstore', \mkont'}
    \stepto^* \tpl{\mval,\mstore'', \mkont'}.
    \hastail(\mtrace,\mkont') \\
  \end{align*}
  \caption{Table invariants}
\label{fig:inv}
\end{figure}
\begin{lemma}[Table invariants]\label{lem:tab-inv}
  If $\inv(\mstate)$ and $\mstate \stepto \mstate'$ then $\inv(\mstate')$.
\end{lemma}

In order to prove this, we need a lemma that allows us to plug in any continuation for memoized results.
%
When traces (sequences of states such that adjacent states are related by $\stepto$) have a common tail, we can replace it with anything.

\begin{lemma}[Stack irrelevance in CESK]\label{lem:irrelevance}
  $\hastail(\mtrace,\mkont)$ implies $\replacetail(\mtrace,\mkont,\mkont')$ is a valid trace.
\end{lemma}
\begin{mathpar}
  \inferrule{ }{\hastail(\epsilon,\mkont)} \quad
  \inferrule{ }{\hastail(\tpl{\mpoint,\mstore,\append{\mkont'}{\mkont}},\mkont)} \quad
  \inferrule{\hastail(\mtrace\mstate,\mkont) \quad
             \mstate \stepto \mstate' \quad
             \hastail(\mstate',\mkont)}
            {\hastail(\mtrace\mstate\mstate',\mkont)}
\end{mathpar}

@@

\begin{claim}(Correctness of PDCFA)\label{thm:pdcfa-tabular}
  Tabular semantics has a bisimulation with the standard semantics with comparable allocation strategies.
\end{claim}

Finally, we can show that even if the tables are shared amongst all states, they are precise enough to not interfere with the results.
\begin{claim}[Equivalence with global tables]\label{thm:global}
  Tabular semantics has a bisimulation with memo-and-k-table-widened semantics.
\end{claim}

@@

@@

This semantics of course enjoys the soundness properties that GC does not introduce dangling pointers (\autoref{thm:gc-pointers}), equal length traces are equal up to GC (in the concrete) (\autoref{thm:gc-concrete}), and abstractions are sound with respect to collected states (\autoref{thm:gc-sound}).
%
Technically since this semantics inspects the touched addresses of the continuation, the set of touched addresses should be considered part of the ``context'' in the $\mathbf{rt}$ continuations.
%
This extra context can be overlooked, since that means the computed touched addresses will be an overapproximation of what is already sound.
%
More touched addresses means a less effective GC, but this is an acceptable tradeoff for a possibly smaller state space.
%
\begin{claim}[No dangling pointers]\label{thm:gc-pointers}
  All addresses in $\Gamma(\mstate)$ are in the domain of $\Gamma(\mstate).\mstore$.
\end{claim}

\begin{claim}[Garbage is irrelevant]\label{thm:gc-concrete}
  If $\mstate \stepto^n \mstate'$ and $\mstate \stepto^n \mstate''$ then $\Gamma(\mstate') = \Gamma(\mstate'')$.
\end{claim}

\begin{claim}[Abstract GC is sound for garbage-free states]\label{thm:gc-sound}
  If $\mstate \stepto \mstate'$ and $\alpha(\Gamma(\mstate)) \sqsubseteq \mastate$ then there is a $\mastate'$ such that $\mastate \astepto \mastate'$ and $\alpha(\Gamma(\mstate')) \sqsubseteq \mastate'$.
\end{claim}

This final proposition is not in the original work on abstract GC; they prove soundness with respect to a concrete semantics that performs GC after every step.
%
The moral correctness of the abstraction is that it never under-approximates \emph{relevant} portions of the state space --- in this case, the reachable subset of the store.
%
The definitions of $\alpha$, $\astepto$ and $\sqsubseteq$ are the simple pointwise abstractions of addresses through states and the reduction relation, and finally a lifted subset relation between states and their components, respectively.
%
Both $\alpha$ and $\sqsubseteq$ are parameterized on the choice of $\Addr$, for which the following soundness criterion should hold:
\begin{mathpar}
  \inferrule*[right={[Sound allocation]}]{\alpha(\mstate) \sqsubseteq \mastate}{\alpha(\alloc(\mstate)) \sqsubseteq \widehat{\alloc}(\mastate)}
\end{mathpar}
\section{Deriving CFA2}
\label{sec:cfa2}

CFA2 is the first published analysis of a higher-order programming language that could properly match calls and returns.
%
We will show that it fits well into the same presentation we gave for PDCFA.
%
Vardoulakis and Shivers had a clear goal of harnessing the extra information a pushdown model provides to produce a high-precision analysis that works well in practice.
%
This resulted in more than just the call/return matching of the previous section, which is why we are showing the two separately.
%

\begin{enumerate}
\item{stack allocation for some bindings in an additional $\mframe \in \Store$}
\item{strong updates on stack frames for resolved non-determinism}
\end{enumerate}
There are two orthogonal features of the semantics, embodied by our semantics' two metafunctions ($\bind$ and $\lookup$, respectively in \autoref{fig:frame-semantics}).
%
The first of these is an addition to the stack-less context.

@@

%%
The second of these is to ameliorate a problem they call ``fake rebinding.''
%
That is, since bindings in the abstract represent several values, we don't want to reference a variable \texttt{x} in two different places and have it resolve to two different values.
%
In AAM, a variable reference non-deterministically steps to all possible values associated with that variable.
%
Here we want to say that once \texttt{x} is considered to stand for value $v$, then all subsequent references of \texttt{x} should be $v$.
%
If they are not all considered $v$, it looks as if \texttt{x} was rebound; it hasn't, and thus it is a ``fake rebinding.''
%
CFA2 does not step to all values on variable reference, but instead carries all its values around in superposition until they need to be observed at, say, a function call.
%
We give a simplified semantics that is more along the AAM style, but CFA2's approach can easily be recovered from it.\footnote{Our Redex model implements the fake rebinding strategy CFA2 itself employs.}
%%

@@

\begin{align*}
  \bind(\mstore,\maddr,\mvar,\mval) &=
   \left\{\begin{array}{ll}
            (\mstore, \snglm{\maddr}{\mval}) &\text{ if } \mvar \text{ allocations never escape} \\
            (\joinone{\mstore}{\maddr}{\mval}, \snglm{\maddr}{\mval}) &\text{ otherwise} \\
          \end{array}\right. \\
  \lookup(\mstore,\mframe,\menv,\mvar,\mlab) &=
    \left\{\begin{array}{ll}
          \setbuild{(\extm{\mframe}{\menv(\mvar)}{\set{\mval}}, \mval)}
                   {\mval \in \mframe(\menv(\mvar))} & \text{if } \mlab \text{ non-escaping} \\
          \setbuild{(\mframe,\mval)}{\mval \in \mstore(\menv(\mvar))} & \text{ otherwise}
           \end{array}\right.
\end{align*}

@@

%%
The stack frame introduces a refinement over the original CESK machine, so we first prove it correct with respect to the concrete semantics.
%
We follow up with a similar proof to that of PDCFA, which is that the CES$\xi$K machine has a bisimulation with its tabular semantics, with comparable allocation strategies.
\begin{claim}[Correctness of stack frame refinement]\label{thm:refinement}
  The CES$\xi$K machine has a bisimulation with the CESK machine with fresh allocation.
\end{claim}
This follows from the invariant that if a stack frame maps $\maddr$ to $\mval$, then the store maps $\maddr$ to $\mval$, meaning that the special lookup function is equivalent to lookup directly in the store (exploits immutability).
%
That is, if we always allocated bindings in the heap.
%
The additional escape information that allows us to elide allocation in the heap introduces a second level of reasoning, where the burden of proof falls on the escape analysis.
%
The additional escape analysis adds information to use in the proof, that if $\mvar$ never escapes, then every reference to allocations of $\mvar$'s binding will always be able to use $\mframe$ for the lookup.
%
With this proposition, we can prove that the CES$\xi$K machine that always allocates in the store has a bisimulation with the refined machine that exploits escape analysis to avoid unnecessary changes to the store.
%

\begin{claim}[Correctness of CFA2]\label{thm:cfa2}
  Tabular semantics has a bisimulation with standard semantics.
\end{claim}
This proof has the same structure as that of PDCFA.

@@

\begin{claim}\label{thm:concrete-sr}
  Tabular semantics has a bisimulation with standard semantics given fresh allocation.
\end{claim}

@@

\begin{claim}\label{thm:sound-sr}
  Tabular semantics simulates standard semantics with abstract allocation.
\end{claim}
